---
title: "Pstat 115, Homework 4"
author: "Marcus Eriksson"
date: "November 9, 2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, 
                      cache=TRUE, 
                      fig.width=5, 
                      fig.height=5,
                      fig.align='center')
indent1 = '    '      
indent2 = paste(rep(indent1, 2), collapse='')
```

###1) Importance Sampling for Galenshore distribution

####a)

For k = 10

```{r}
#create galenshore distribution
library(tidyverse)
library(ggplot2)


galen <- function(x, alpha, theta){
  
  (2/ gamma(alpha))*(theta^(2*alpha))*x^(2*alpha-1)*exp((-theta^2)*x^2)
}


set.seed(1)
#for N = 10
alpha = 1
theta = .5
q = function(x) dexp(1)
p = function(x) galen(x, alpha, theta)
N_1a = 10
Ns_1a = 1:N_1a
estimate = rep(0, N_1a)
for(n in Ns_1a) {
x = rexp(n)
w = p(x) / q(x)
estimate[n] = sum(x * w) / sum(w)
}

list(Ns_1a = Ns_1a,
estimate = estimate) %>%
as.tibble %>%
ggplot(aes(Ns_1a, estimate)) +
geom_point() +
geom_hline(aes(yintercept = gamma(alpha+.5) / theta * gamma(alpha)), color = "red") +
 
ggtitle("Estimates for E[x] varying numbers of samples in black\nTruth in red, for N= 10")


```

For k = 100

```{r}
#for N = 100
set.seed(1)
N_100a = 100
Ns_100a = 1:N_100a
estimate = rep(0, N_100a)
for(n in Ns_100a) {
x = rexp(n)
w = p(x) / q(x)
estimate[n] = sum(x * w) / sum(w)
}

list(Ns_100a = Ns_100a,
estimate = estimate) %>%
as.tibble %>%
ggplot(aes(Ns_100a, estimate)) +
geom_point() +
geom_hline(aes(yintercept = gamma(alpha+.5) / theta * gamma(alpha)), color = "red") +
ggtitle("Estimates for E[x] varying numbers of samples in black\nTruth in red, for N= 100")

```

For k = 1000
```{r}
set.seed(13)
#for N = 1000
N_1000a = 1000
Ns_1000a = 1:N_1000a
estimate = rep(0, N_1000a)
for(n in Ns_1000a) {
x = rexp(n)
w = p(x) / q(x)
estimate[n] = sum(x * w) / sum(w)
}
estimate[n]

list(Ns_1000a = Ns_1000a,
estimate = estimate) %>%
as.tibble %>%
ggplot(aes(Ns_1000a, estimate)) +
geom_point() +
geom_hline(aes(yintercept = gamma(alpha+.5) / theta * gamma(alpha)), color = "red") +
ggtitle("Estimates for E[x] varying numbers of samples in black\nTruth in red, for N= 1000")

```


####b) Histogram of log weights for k=1000

```{r}

hist(log(w))
```


####c) Sampling from a normal distribution:

For k = 10
```{r}
set.seed(1)
alpha = 1
theta = .5
q = function(x) dnorm(x, 1.5 ,1 )
p = function(x) galen(x, alpha, theta)
N_b10 = 10
Ns_b10 = 1:N_b10
estimate = rep(0, N_b10)
for(n in Ns_b10) {
x = rnorm(n, 1.5, 1)
w = p(x) / q(x)
estimate[n] = sum(x * w) / sum(w)
}
estimate[n]

list(Ns_b10 = Ns_b10,
estimate = estimate) %>%
as.tibble %>%
ggplot(aes(Ns_b10, estimate)) +
geom_point() +
geom_hline(aes(yintercept = gamma(alpha+.5) / theta * gamma(alpha)), color = "red") +
ggtitle("Estimates for E[x] varying numbers of samples in black\nTruth in red, for N= 10")
```


For k = 100

```{r}
#n = 100
set.seed(1)
alpha = 1
theta = .5
q = function(x) dnorm(x, 1.5, 1)
p = function(x) galen(x, alpha, theta)
N_b100 = 100
Ns_b100 = 1:N_b100
estimate = rep(0, N_b100)
for(n in Ns_b100) {
x = rnorm(n, 1.5, 1)
w = p(x) / q(x)
estimate[n] = sum(x * w) / sum(w)
}
estimate[n]

list(Ns_b100 = Ns_b100,
estimate = estimate) %>%
as.tibble %>%
ggplot(aes(Ns_b100, estimate)) +
geom_point() +
geom_hline(aes(yintercept = gamma(alpha+.5) / theta * gamma(alpha)), color = "red") +
ggtitle("Estimates for E[x] varying numbers of samples in black\nTruth in red, for N= 100")
```

For k = 1000

```{r}
#n = 1000

set.seed(1)
alpha = 1
theta = .5
q = function(x) dnorm(x, 1.5, 1)
p = function(x) galen(x, alpha, theta)
N_b1000 = 1000
Ns_b1000 = 1:N_b1000
estimate = rep(0, N_b1000)
for(n in Ns_b1000) {
x = rnorm(n, 1.5, 1)
w = p(x) / q(x)
estimate[n] = sum(x * w) / sum(w)
}
estimate[n]

list(Ns_b1000 = Ns_b1000,
estimate = estimate) %>%
as.tibble %>%
ggplot(aes(Ns_b1000, estimate)) +
geom_point() +
geom_hline(aes(yintercept = gamma(alpha+.5) / theta * gamma(alpha)), color = "red") +
ggtitle("Estimates for E[x] varying numbers of samples in black\nTruth in red, for N= 1000")

#print true mean:
gamma(alpha+.5) / theta * gamma(alpha)

```

Histogram of log importance weights for k = 1000

```{r}
hist(log(w))
```



The true mean of our Galenshore distribution is 1.772454; our k = 1000 estimate we obtained while using the exponential distribution for importance sampling was about 1.12 compared to our estimate from the normal distribution which was much further away from the true mean at about 7.54. It is without question better to use the proposal exp(1) to perform importance sampling for the Galenshore distribtion we were given. The histograms of the weights also show that the exponential distribution produces better samples.


###2) Rejection sampling for the semi - circle distribution:


####a) Identify a proposal distribution for rejection sampling:

Proposal distribution: Uniform distribution with mean 1


####b) Plot the semi - circle distribution for R = 1

```{r}
#semi circle distribution:
semi_circle <- function(x, R){
  (2/(pi*R^2))*sqrt((R^2 - (x^2)))
  
}


R = 1
N = 1000
mode = 0
maxB = 1
list(x = runif(N),
y = runif(N) * maxB) %>%
as.tibble %>%
mutate(use = y < semi_circle(x, R)) %>%
ggplot(aes(x, y)) +
geom_point(aes(color = use)) +
stat_function(fun = function(x) semi_circle(x, R)) +
geom_rect(aes(xmin = 0.0, xmax = 1.0, ymin = 0.0, ymax = maxB), alpha = 0.0, color = "red") +
ggtitle("Proposal distribution outlined in red, target in black")
```


####c) Perform rejection sampling for R =  .1, .2....3.0

```{r}
R_1 <- seq(from = .1 , by =.1, to = 3)

N = 1000
mode = 0
maxB = 1
list(x = runif(N),
y = runif(N) * maxB) %>%
as.tibble %>%
mutate(use = y < semi_circle(x, R_1)) %>%
ggplot(aes(x, y)) +
geom_point(aes(color = use)) +
stat_function(fun = function(x) semi_circle(x, R_1)) +
geom_rect(aes(xmin = 0.0, xmax = 1.0, ymin = 0.0, ymax = maxB), alpha = 0.0, color = "red") +
ggtitle("Proposal distribution outlined in red, target in black")

```


###3) Rejection sampling and interval estimation:

####a) Use rejection sampling to sample from p(x) = |sin(x)|*I{0, 2pi}


```{r}

#install.packages("HDInterval")
#create indicator function
library(HDInterval)

ivec <- 0+ ( 0 < x & x < 2*pi)

set.seed(1)

sin_dist <- function(x){
  abs(sin(x)) * ivec
}

R = 1
N_3a = 1000

maxB = 1
list(x = runif(N_3a),
y = runif(N_3a) * maxB) %>%
as.tibble %>%
mutate(use = y < sin_dist(x)) %>%
ggplot(aes(x, y)) +
geom_point(aes(color = use)) +
stat_function(fun = function(x) sin_dist(x)) +
geom_rect(aes(xmin = 0.0, xmax = 1.0, ymin = 0.0, ymax = maxB), alpha = 0.0, color = "red") +
ggtitle("Proposal distribution outlined in red, target in black")


#compute middle 50%
quantile(sin_dist(x), .25)
quantile(sin_dist(x), .75)

#HPDI 
hdi(sin_dist(x), credMass = .5, allowSplit = TRUE)


 

```

50% Quantile Interval: (.4303093, 0.8340892) 
Length: 0.4037799

50% Highest Posterior Density Interval: (0.4918161, 0.9060101)
Length: 0.414194

The HPD interval is slightly larger than the quantile interval;  


####b) Plot p(x)


```{r}
curve(sin_dist(x), main = "Red is HPD interval and Blue is Quantile interval ")
abline(v =.4303093, col = "Blue")
abline(v =.8340892, col = "Blue")

abline(v =.4918161, col = "Red")
abline(v =.9060101, col = "Red")
```







