---
title: "Homework 3 PSTAT 115, Fall 2018"
author: "Eddie Bermudez & Marcus Eriksson"
date: "__Due on October 28, 2018 at 11:59 pm__"
output: pdf_document
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo=TRUE, 
                      cache=FALSE, 
                      fig.width=5, 
                      fig.height=5,
                      fig.align='center')
indent1 = '    '
indent2 = paste(rep(indent1, 2), collapse='')
indent3 = paste(rep(indent1, 3), collapse='')
r = function(x, digits=2){ round(x, digits=digits) }
library(tidyverse)
library(reshape2)
```

## 1(a)

```{r}
y.A <- c(12,9,12,14,13,13,15,8,15,6)
y.B <- c(11,11,10,9,9,8,7,10,6,8,8,9,7)

#sum.A <- sum(y.A)
sum.A <- 117
sum.B <- 113
#sum.B <- sum(y.B)

alpha.A <- 120
beta.A <- 10
n.A <- 10
#n.A <- length(y.A)

n.0 <- 1:50

#alpha.B=12
#beta.B=1
alpha.B=12*n.0
beta.B=n.0
n.B <- 13
#n.B <- length(y.B)

set.seed(99)
mc.theta.A <- rgamma(10000, alpha.A+sum.A, beta.A+n.A)
mc.theta.B <- rgamma(10000, alpha.B+sum.B, beta.B+n.B)
mc.y.A <- rpois(10000, mc.theta.A)
mc.y.B <- rpois(10000, mc.theta.B)

#Probability
mean(mc.theta.B<mc.theta.A)
mean(mc.y.B<mc.y.A)

#Posterior Distribution
theta <- seq(0,20,length=10000)
plot(theta, dgamma(theta,alpha.A+sum.A,beta.A+n.A), 
     type = "l", col = "red",
     main = "Posterior distribution",
     xlab = expression(theta),
     ylab = expression(p(theta(y[1],...,y[n])))
     )
lines(theta, dgamma(theta,alpha.B+sum.B,beta.B+n.B), col = "blue")
legend("topright", legend = c("red=group A", "blue=group B"), bty="n")


#Line Plot
set.seed(99)
for(i in 1:50){
  alpha.B <- 12*i
  beta.B <- i
  theta.A <- rgamma(10000, alpha.A+sum.A, beta.A+n.A)
  theta.B <- rgamma(10000, alpha.B+sum.B, beta.B+n.B)
  #n.0[i] <- mean(theta.B<theta.A)
  
  mc.y.A <- rpois(10000, theta.A)
  mc.y.B <- rpois(10000, theta.B)
  n.0[i] <- mean(mc.y.B<mc.y.A)
}
plot(n.0, type = "l",
     main = "1(a) line plot",
     xlab = "n_0",
     ylab = "Pr(theta_B < theta_A | y_A, y_B)")
```
 
 Pr(theta_B < theta_A | y_A, y_B) for a range of values of n_0 is about 0.5538. The posterior distribution appears to be sensitive and depends on the prior distribution.
 
 

## 1(b)
    
```{r}
set.seed(7)
mc.tY.A <- rgamma(10000, alpha.A+sum.A, beta.A+n.A)
mc.tY.B <- rgamma(10000, alpha.B+sum.B, beta.B+n.B)
mc.y2.A <- rpois(10000, mc.tY.A)
mc.y2.B <- rpois(10000, mc.tY.B)

mean(mc.tY.B<mc.tY.A)
mean(mc.y2.B<mc.y2.A)

set.seed(7)
for(i in 1:50){
  alpha.B <- 12*i
  beta.B <- i
  tY.A <- rgamma(10000, alpha.A+sum.A, beta.A+n.A)
  tY.B <- rgamma(10000, alpha.B+sum.B, beta.B+n.B)
  #n.0[i] <- mean(tY.B<tY.A)
  
  mc.y2.A <- rpois(10000, tY.A)
  mc.y2.B <- rpois(10000, tY.B)
  n.0[i] <- mean(mc.y2.B<mc.y2.A)
}
plot(n.0, type = "l",
     main = "1(b) line plot",
     xlab = "n_0",
     ylab = "Pr(tildaY_B < tildaY_A | y_A, y_B)")
```
 Pr(~YB < ~YA | y_A, y_B) for a range of values of n_0 is about 0.5013. The posterior distribution appears to be sensitive and depends on the prior distribution.
 
## 1(c)

The posterior distribution of events {thetaB < thetaA} don't seem to be affected after replacement with event {~YB < ~YA}.

## 2(a)

```{r}
y.A <- c(12,9,12,14,13,13,15,8,15,6)
sum.A <- 117
#sum.A <- sum(y.A)

alpha.A <- 120
beta.A <- 10
n.A <- 10
#n.A <- length(y.A)

#for dataset y_A
mean.yA <- mean(y.A)
mean.yA
v.yA <- var(y.A)*(length(y.A)-1)/length(y.A)
v.yA
mean.yA/v.yA

#for theta_A
E.A <- (alpha.A+sum.A)/(beta.A+n.A)
V.A <- (alpha.A+sum.A)/(beta.A+n.A)^2
E.A/V.A

y.B <- c(11,11,10,9,9,8,7,10,6,8,8,9,7)
sum.B <- 113
#sum.B <- sum(y.B)

alpha.B=12
beta.B=1
n.B <- 13
#n.B <- length(y.B)

#for dataset y_B
mean.yB <- mean(y.B)
mean.yB
v.yB <- var(y.B)*(length(y.B)-1)/length(y.B)
v.yB
mean.yB/v.yB

#for thetha_B
E.B <- (alpha.B+sum.B)/(beta.B+n.B)
V.B <- (alpha.B+sum.B)/(beta.B+n.B)^2
E.B/V.B

```

If the Poisson model were a reasonable one, a typical value of t^(s) would be the sample average of a poisson divided by the sample variance of a poisson. For a poisson model with parameter lambda, the expectation or mean is lambda and the variance is lambda. lambda/lambda is 1. 

## 2(b)

```{r}
#histogram

set.seed(50)
#mc.theta.A <- rgamma(1000, alpha.A+sum.A, beta.A+n.A)
#mc.y.A <- rpois(1000, mc.theta.A)
tsA<- NULL
for(i in 1:1000){
  theta.A <- rgamma(10, alpha.A+sum.A, beta.A+n.A)
  mc.y.A <- rpois(10, theta.A)
  #n.0[i] <- mean(theta.B<theta.A)
  tsA[i] <- (mean(mc.y.A)/var(mc.y.A))
}

hist(tsA,
     main = "Histogram of t^(s) for y_A",
     xlab = "t^(s)")
abline(v=1.391201, col = "Lime Green")
```

The poisson model seems reasonable for this dataset because the mean/variance was about 1.4 for the data and the histogram seems to reflect this well.


## 2(c)

```{r}
#repeat 2b

set.seed(51)
#mc.theta.B <- rgamma(1000, alpha.B+sum.B, beta.B+n.B)
#mc.y.B <- rpois(1000, mc.theta.B)
tsB<- NULL
for(i in 1:1000){
  theta.B <- rgamma(13, alpha.B+sum.B, beta.B+n.B)
  mc.y.B <- rpois(13, theta.B)
  #n.0[i] <- mean(theta.B<theta.A)
  tsB[i] <- (mean(mc.y.B)/var(mc.y.B))
}
hist(tsB,
     main = "Histogram of t^(s) for y_B",
     xlab = "t^(s)")
abline(v=3.927807, col = "Purple")
```

The poisson model does not seem to fit this data that well because of the fact that the mean/variance is about 3.9 and the simulated data from a poisson distribution does not match this since it produced data centered around 1 as seen in this histogram.
