---
title: "Pstat 127 Homework 2"
author: "Marcus Eriksson"
date: "Section: W 5pm"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



###1) There are 681 cases of potentially cancerous tumors of which 238 are actually malignant. Determining whether a tumor is really malignant is traditionally determined by an invasive surgical procedure. The purpose of this study was to determine whether a new procedure called fine needle aspiration which draws only a small sample of tissue could be effective in determining tumor status.

####a) Binomial regression with class as the response and other variables as predictors:

```{r}
library(faraway)
data("wbca")
#wbca

wbca_glm <- glm(Class~. , family = "binomial", data = wbca)
summary(wbca_glm)
```

Residual deviance: 89.464 with 671 degrees of freedom.

####b) What does the model from part a) assume for the dispersion parameter phi?


```{r}
sum(residuals(wbca_glm, type = "deviance")^2)/(691-10)
```

The model from part a) assumes a dispersion parameter of 1 since it is a binomial model. When estimating the dispersion parameter we get 0.1313718, based on this estimate the assumed value of 1 for phi does seem possible.


####c) Using backwards selection to select a subset of predictors:

```{r}
step(wbca_glm, direction = "backward")

reduced_glm <- glm(Class~ Adhes + BNucl + Chrom + Mitos + NNucl + 
    Thick + UShap, family = "binomial", data = wbca)
#reduced_glm
```

####d) Using the reduced model from part c), estimate log odds and probability that a specific tumor is benign:


```{r}

cancer <- c(Adhes = 1, BNucl = 1, Chrom = 3, Epith = 2, Mitos = 1, NNucl = 1, Thick = 4,
UShap = 1, USize = 1)


                 
cancer_df <- data.frame(matrix(cancer,nrow=1))
#is.data.frame(cancer_df)
#cancer_df
colnames(cancer_df) <- c("Adhes", "BNucl", "Chrom", "Epith", "Mitos", "NNucl", "Thick",
"UShap", "USize")
#cancer_df

predict.glm(reduced_glm, newdata = cancer_df  ,type = "link") #log odds
predict.glm(reduced_glm, newdata = cancer_df  ,type = "response") #probability
```

Log-odds: 4.834
Probability: .9921

####e) data from another tumor with same predictor values as in d) but Chrom = 4:


```{r}
cancer <- c(Adhes = 1, BNucl = 1, Chrom = 4, Epith = 2, Mitos = 1, NNucl = 1, Thick = 4,
UShap = 1, USize = 1)
cancer_df <- data.frame(matrix(cancer,nrow=1))
#is.data.frame(cancer_df)
#cancer_df
colnames(cancer_df) <- c("Adhes", "BNucl", "Chrom", "Epith", "Mitos", "NNucl", "Thick",
"UShap", "USize")
#cancer_df


predict.glm(reduced_glm, newdata = cancer_df  ,type = "link") #log odds
```
The log-odds of tumor B being benign is lower than the log-odds of tumor A being benign.The log odds are lower by a about .57


####f) Tumor is classified as benign if p > .5 and malignant if p < .5; compute # of errors of both types if applied to reduced model:


```{r}
predict_wbca <- predict.glm(reduced_glm, type = "response")
table(predict_wbca < .5, wbca$Class)
```

227 predicted to be malignant when in reality there are 238 malignant tumors.
434 predicted to be benign when in reality there are 443 benign cases. 
We have 20 total errors for a misclassification rate of .0293.










