---
title: "Pstat 115, Homework 6, Problem 1"
author: "Marcus Eriksson"
date: "Section W 4pm"
output: pdf_document
---

###1) Logistic regression for toxicity data:

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, 
                      cache=TRUE, 
                      fig.width=5, 
                      fig.height=5,
                      fig.align='center')
indent1 = '    '      
indent2 = paste(rep(indent1, 2), collapse='')
```




```{r}
x <- c(-0.86, -0.3, -0.05, 0.73)
n <- rep(5, 4)
y <- c(0, 1, 3, 5)

```


####a) See attached file


####b) Solve for LD50:

```{r}
dbinom(y, n, .5, log = FALSE)
```



####c) Implementing a Metropolis Hastings algorithm:

```{r}
set.seed(1)
library(coda)
library(MASS)
library(tidyverse)
alpha_t = 1
beta_t = 1

mvrnorm(1, c(alpha_t, beta_t), matrix(c(1.0, 0.0, 0.0, 1.0), nrow = 2))


logp = function(theta) {
sum(dnorm(theta, c(1.0, 1.0), c(1.0, 1.5), log = TRUE))
}
metropolis = function(theta_s, cov) {
# Note: theta_s is a vector now!
theta_p = mvrnorm(1, theta_s, cov)

if(logp(theta_p) > logp(theta_s)) {
return(theta_p)
} else {
r = runif(1)
if(log(r) < logp(theta_p) - logp(theta_s)) {
return(theta_p)
} else {
return(theta_s)
}
}
}
# theta_0 is your initial parameter guess
# burnin is number of burnin samples
# maxit are the number of samples you generate (post burnin)
# cov is the proposal covariance matrix
rw_metrop_multi = function(theta_0, burnin, maxit, cov) {
samples = matrix(0, ncol = length(theta_0), nrow = burnin + maxit)
samples[1,] = theta_0
for(i in 1:(nrow(samples) - 1)) {
samples[i + 1,] = metropolis(samples[i,], cov)
}
samples[burnin:(burnin + maxit),]
}


samples = rw_metrop_multi(c(10.0, 10.0), 1000, 10000, matrix(c(1.0, 0.0, 0.0, 1.0), nrow = 2))

plot(as.mcmc(samples))
effectiveSize(samples)
acf(samples)
```
Effective sample size and acceptance rates: 
for alpha: 1110.7642, 11.1%
for beta: 597.1462, 5.97%


####d) Tune the parameters to achieve a reasonable acceptance rate:


```{r}
set.seed(1)
alpha_t = 1
beta_t = 1

mvrnorm(1, c(alpha_t, beta_t), matrix(c(1.0, 0.0, 0.0, 1.0), nrow = 2))


logp = function(theta) {
sum(dnorm(theta, c(1.0, 1.0), c(1.0, 1.5), log = TRUE))
}
metropolis = function(theta_s, cov) {
# Note: theta_s is a vector now!
theta_p = mvrnorm(1, theta_s, cov)

if(logp(theta_p) > logp(theta_s)) {
return(theta_p)
} else {
r = runif(1)
if(log(r) < logp(theta_p) - logp(theta_s)) {
return(theta_p)
} else {
return(theta_s)
}
}
}
# theta_0 is your initial parameter guess
# burnin is number of burnin samples
# maxit are the number of samples you generate (post burnin)
# cov is the proposal covariance matrix
rw_metrop_multi = function(theta_0, burnin, maxit, cov) {
samples = matrix(0, ncol = length(theta_0), nrow = burnin + maxit)
samples[1,] = theta_0
for(i in 1:(nrow(samples) - 1)) {
samples[i + 1,] = metropolis(samples[i,], cov)
}
samples[burnin:(burnin + maxit),]
}


samples = rw_metrop_multi(c(10.0, 10.0), 100, 10000, matrix(c(10, .0, .00, 10), nrow = 2))

plot(as.mcmc(samples))
effectiveSize(samples)
acf(samples)

```






####e) Using output from the Metropolis algorithm, make a plot showing 50% & 95% confidence bands for the probability of death:

50% and 95% confidence bands:

```{r}

xgrid <- seq(-1, 1, by = 0.01)


compute_curve <- function(samples) {
beta_0 <- samples[1]
beta_1 <- samples[2]
beta_0 + beta_1 * xgrid
}

res <- apply(samples, 1, compute_curve)
quantiles <- apply(res, 1, function(x) quantile(x, c(0.025, 0.25, 0.75, 0.975)))
posterior_mean <- rowMeans(res)
tibble(x=xgrid, q025=quantiles[1, ], q25=quantiles[2, ],
q75=quantiles[3,], q975=quantiles[4, ], mean=posterior_mean) %>% ggplot() +
geom_ribbon(aes(x=xgrid, ymin=q025, ymax=q975), alpha=0.2) +
geom_ribbon(aes(x=xgrid, ymin=q25, ymax=q75), alpha=0.5) +
geom_line(aes(x=xgrid, y=posterior_mean), size=1)
# geom_vline(xintercept = mean(-samples[, 1]/samples[, 2]), linetype="dashed") +
# geom_hline(yintercept = 0.5, linetype="dashed")

ggplot(tibble(x=xgrid, q025=quantiles[1, ], q25=quantiles[2, ],
q75=quantiles[3,], q975=quantiles[4, ],mean=posterior_mean)) +
geom_ribbon(aes(x=xgrid, ymin=q025, ymax=q975), alpha=0.2) +
geom_ribbon(aes(x=xgrid, ymin=q25, ymax=q75), alpha=0.5) +
geom_line(aes(x=xgrid, y=posterior_mean), size=1) +
geom_point(data =data.frame(x= x, y = y) , aes(x= x, y = y), col = "blue")+ggtitle("50% and 95% confidence band for probability of death")
```


