---
title: "PSTAT 131, HW3"
author: "Marcus Eriksson"
date: "November 12, 2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, 
                      cache=TRUE, 
                      fig.width=5, 
                      fig.height=5,
                      fig.align='center')
indent1 = '    '      
indent2 = paste(rep(indent1, 2), collapse='')
```




```{r}
#install.packages("dendextend")
#install.packages("NbClust")
#install.packages("ggridges")
#install.packages("superheat")
#install.packages("tree")
library(tidyverse)
library(ROCR)
library(tree)
library(maptree)
library(class)
library(lattice)
library(ggridges)
library(superheat)
```


```{r}
drug_use <- read_csv('drug.csv',
col_names = c('ID','Age','Gender','Education','Country','Ethnicity',
'Nscore','Escore','Oscore','Ascore','Cscore','Impulsive',
'SS','Alcohol','Amphet','Amyl','Benzos','Caff','Cannabis',
'Choc','Coke','Crack','Ecstasy','Heroin','Ketamine',
'Legalh','LSD','Meth','Mushrooms','Nicotine','Semer','VSA'))
```

###1)

```{r}
drug_use <- drug_use %>% mutate_at(as.ordered, .vars=vars(Alcohol:VSA))
drug_use <- drug_use %>%
mutate(Gender = factor(Gender, labels=c("Male", "Female"))) %>%
mutate(Ethnicity = factor(Ethnicity, labels=c("Black", "Asian", "White",
"Mixed:White/Black", "Other",
"Mixed:White/Asian",
"Mixed:Black/Asian"))) %>%
mutate(Country = factor(Country, labels=c("Australia", "Canada", "New Zealand",
"Other", "Ireland", "UK", "USA")))
```


####a)  Define a new factor response variable 


```{r}
drug_use <- drug_use %>%
mutate(recent_cannabis_use=factor(ifelse(Cannabis >= "CL3", "Yes", "No"),
levels=c("No", "Yes")))

```



####b) We will create a new tibble that includes a subset of the original variables. 

```{r}
drug_use_subset <- drug_use %>% select(Age:SS, recent_cannabis_use)
```



Split drug_use_subset into a training data set and a test data set called drug_use_train and drug_use_test.
The training data should include 1500 randomly sampled observation and the test data should include the
remaining observations in drug_use_subset. Verify that the data sets are of the right size by printing
dim(drug_use_train) and dim(drug_use_test)


```{r}
train_index <- sample(nrow(drug_use_subset), 1500)
drug_use_train <- drug_use_subset[train_index,]
drug_use_test <- drug_use_subset[-train_index, ]
```


####c) Fit a logistic regression to model recent_cannabis_use

```{r}
drug.lm <- glm(drug_use_train$recent_cannabis_use ~. , data = drug_use_train, family = binomial)

summary(drug.lm)
```




####2) Decision tree models of drug use:


```{r}
tree_parameters = tree.control(nobs=nrow(drug_use_train), minsize=10, mindev=1e-3)
```

This sets the smallest number of allowed observations in each leaf node to 10 and requires a deviance of at
least 1e-3 to split a node.

####a) (a). Use 10-fold CV to select the a tree which minimizes the cross-validation misclassification rate.

```{r}
set.seed(1)

drug_tree <- tree(drug_use_train$recent_cannabis_use ~., data = drug_use_train)

cv.tree(drug_tree , FUN = prune.misclass, K = 10)

best_size = 6

```


####b) . Prune the tree to the size found in the previous part and plot the tree using the draw.tree function


```{r}
drug_tree.pruned <- prune.tree(drug_tree, k = best_size)

draw.tree(drug_tree.pruned, nodeinfo = TRUE)
```

The first variable to be split in our tree is country.


####c) Compute and print the confusion matrix for the test data 

```{r}
drug.truth <- drug_use_test$recent_cannabis_use 

drug.pred <- predict(drug_tree.pruned, drug_use_test, type = "class")

error = table(drug.pred, drug.truth)
error


#TPR = TP/(TP +FP)

148/(148+49)

#FPR = FP/(FP+TN)

27/(27+161)
```

TPR = 0.751269
FPR = 0.143617

###3) Model Comparison:

####a) . Plot the ROC curves for both the logistic regression fit and the decision tree on the same plot.

```{r}
library(ROCR)
prob_tree_drug <- predict(drug_tree.pruned, drug_use_test, type="vector")

prob_lm_drug <- predict(drug.lm, drug_use_test, type = "response")


predict_tree_drug <- prediction(prob_tree_drug[,2], drug_use_test$recent_cannabis_use)
predict_logistic_drug <- prediction(prob_lm_drug, drug_use_test$recent_cannabis_use)

perf_tree_drug = performance(predict_tree_drug, measure = "tpr", x.measure = "fpr")
perf_logistic_drug = performance(predict_logistic_drug, measure = "tpr", x.measure = "fpr")

plot(perf_tree_drug@x.values[[1]], perf_tree_drug@y.values[[1]], type = "l", col = "Light Blue", lwd =2, main = "ROC curve for Tree Model")
abline(0,1)
lines(perf_logistic_drug@x.values[[1]], perf_logistic_drug@y.values[[1]], type = "l", col = "Dark Blue", lwd = 2, main = "ROC curve for logistic Regression Model")



```


####b) Compute the AUC for both models and print them. Which model has larger AUC?

```{r}
auc_tree = performance(predict_tree_drug, "auc")@y.values
auc_tree

auc_logistic = performance(predict_logistic_drug, "auc")@y.values
auc_logistic
```


The logistic regression model has a higher AUC; it has an AUC of 0.8931496 compared to the AUC of the tree model of 0.8501112


###4)  Clustering and dimension reduction for gene expression data

```{r}
leukemia_data <- read_csv("leukemia_data.csv")
leukemia_data

```

####a) Convert the Type column to a factor using the mutate function. 



```{r}
leukemia_data <- leukemia_data %>% 
        mutate(Type = as.factor(Type))

table(leukemia_data$Type)

class(leukemia_data$Type)
```

The leukemia type that occurs the least in our dataset is: BCR-ABL


####b) Run PCA on the leukemia data using prcomp function with scale=TRUE and center=TRUE 


```{r}

leukemia_sub <- prcomp(leukemia_data[2:3142], scale = TRUE, center = TRUE)

sdev <- leukemia_sub$sdev

pve <- (sdev^2) / sum(sdev^2)

cumulative_pve <-  cumsum(pve)



par(mfrow=c(1, 2))

plot(pve, type="l", lwd=3)
plot(cumulative_pve, type="l", lwd=3)

```



####c) Use the results of PCA to project the data into the first two principal component dimensions. 

```{r}
rainbow_colors <- rainbow(7)
plot_colors <- rainbow_colors[leukemia_data$Type]

leukemia_label <- leukemia_sub$x[, 1:2]

plot(leukemia_label, xlim=c(-60, 60), ylim=c(-60, 60), cex=.5)
text( leukemia_label, labels = leukemia_data$Type, col = plot_colors, cex = .6)

pc1_vector <- leukemia_sub$rotation[,1]

pc1_vector= abs(pc1_vector)

pc1_vector = sort(pc1_vector, decreasing = TRUE)
head(pc1_vector)
```

Along the PC1 axis T-ALL appears the most seperated from rest of the data points.


####f) Compute a euclidean distance matrix between the subjects using the dist function and then run hierarchical clustering using complete linkage. Plot two dendrograms based on the hierarchical clustering result.

```{r}

leukemia_subset <- filter(leukemia_data, Type == "T-ALL" | Type == " TEL-AML1" | Type == "Hyperdip50")

#leukemia_subset
library(NbClust)
leukemia_dist <- dist(leukemia_subset, method="euclidean")


leukemia.hc = hclust(leukemia_dist, method="complete")
leukemia.hc

library(dendextend)
## dendrogram: branches colored by 3 groups
dend1 = as.dendrogram(leukemia.hc)
# color branches and labels by 3 clusters
dend1 = color_branches(dend1, k=3)
dend1 = color_labels(dend1, k=3)
# change label size
dend1 = set(dend1, "labels_cex", 0.3)

dend1 = set_labels(dend1, labels=leukemia_subset$Type[order.dendrogram(dend1)])
# plot the dendrogram
plot(dend1, horiz=T, main = "Dendrogram colored by three clusters")

#second dendrogram
dend2 = as.dendrogram(leukemia.hc)
# color branches and labels by 3 clusters
dend2 = color_branches(dend2, k=5)
dend2 = color_labels(dend1, k=5)
# change label size
dend2 = set(dend2, "labels_cex", 0.3)

dend2 = set_labels(dend2, labels=leukemia_subset$Type[order.dendrogram(dend2)])
# plot the dendrogram
plot(dend2, horiz=T, main = "Dendrogram colored by five clusters")

```












